\input{1-intro.ltx} 

\input{2-case-study.tex}

\input{3-plain-xtext.tex}

\input{4-xbase.tex}

\section{Xsemantics}

Lorenzo Bettini's type system Xsemantics, formerly called XtypeS~\cite{Bet11},
uses an Xtext metamodel and a set of rules and premises as input and returns a set of generated Java methods that can be used in Xtext's scoping and validator classes.
\subsection{Type System Specification}
Xsemantics offers a declarative specification of type rules by using
Judgements
rules declaratively specify statements that must hold for the type system, example

checking rules

axioms

predefined operations

\subsection{Usage in Xtext}

With the declarative specification (written in the Xsemantics DSL), Java methods are generated that can be used in the SopeProvider or Validators. 

If the declarative rules are marked with […], Xtext validator classes are directly created and no further work is necessary.

Example output for reusable methods

\begin{lstlisting}
TypeSystemResult<MyOtherGrammarElement> typeAsMyOtherGrammarElement(MyGrammarElement left); 
TypeSystemResult<MyGrammarElement, MyOtherGrammarElement> type(MyGrammarElement left, MyOtherGrammarElement right); 
TypeSystemResult<Boolean> checkType(MyGrammarElement left, MyOtherGrammarElement right); 
\end{lstlisting}

\section{Xtext Type System}

Markus Völter's type system allows declarative and procedural type constraint specification. It features a Java class called DefaultTypesystem which implements Itypesystem and provides methods that can be used in the declarative specification of the developer's type system rules. In the future, a DSL may be created for the declarative type specification. A DSL for the API is also provided.

\subsection{Type System Specification}

\subsection{Usage in Xtext}

@Inject 
private ITypesystem ts; 

Suggests to insert type information into a global index entry (IEObjectDescription)
Case Study
Featherweight Java?
Domain Model / LWC?
Case Study with plain Xtext
Case Study with Xbase
Case Study with Xtext Type System
Case Study with Xsemantics
\section{Comparison}
Comparison
Comparison Criteria
Table

Preis
Installation
Dokumentation
Lines of Code added nach Sprache (Java, Xtend, DSL?)
Support

\section{Conclusion}
